Psychological Needs During Intergroup Contact: Three Experience Sampling Studies

### 2022-11-11 TitlePage.tex
shortened + bib fixes

\vspace{1.5em}
\noindent\textit{Open Science Practices:}\\
\noindent \href{https://osf.io}{\includegraphics[height=\badgeheight]{assets/open-badges-small/registrationplus-color.png}} Preregistration+, 
\href{https://osf.io}{\includegraphics[height=\badgeheight]{assets/open-badges-small/material-color.png}} Open Materials, 
\href{https://osf.io}{\includegraphics[height=\badgeheight]{assets/open-badges-small/data-color.png}} Open Data,
\href{https://osf.io}{\includegraphics[height=\badgeheight]{assets/open-badges-small/code-color.png}} Open Code, 
\href{https://osf.io}{\includegraphics[height=\badgeheight]{assets/open-badges-small/supplements-color.png}} Open Supplements

>>>>

\vspace{1.5em}
\noindent\textit{Open Science Practices:}\\
\noindent \href{https://doi.org/10.17605/OSF.IO/R5E8C}{\includegraphics[height=\badgeheight]{assets/open-badges-small/registrationplus-color.png}} Preregistration+, 
\href{https://osf.io/pr9zs/?view_only=208a53a1f0ff48dda1c17357328fa578}{\includegraphics[height=\badgeheight]{assets/open-badges-small/material-color.png}} Open Materials, 
\href{https://osf.io/pr9zs/?view_only=208a53a1f0ff48dda1c17357328fa578}{\includegraphics[height=\badgeheight]{assets/open-badges-small/data-color.png}} Open Data,
\href{https://osf.io/pr9zs/?view_only=208a53a1f0ff48dda1c17357328fa578}{\includegraphics[height=\badgeheight]{assets/open-badges-small/code-color.png}} Open Code, 
\href{https://osf.io/pr9zs/?view_only=208a53a1f0ff48dda1c17357328fa578}{\includegraphics[height=\badgeheight]{assets/open-badges-small/supplements-color.png}} Open Supplements

### 2022-11-11 manuscript.tex
shortened + bib fixes

% Abstract
\abstract{
One challenge of modern intergroup contact research has been the question of when and why an interaction is perceived as positive and leads to better intergroup relations. We propose to consider situational psychological needs during everyday intergroup contact. We conducted three extensive longitudinal studies with recent migrants, to capture their interactions with the majority outgroup (total N of measurements = 10,297). The novel psychological needs mechanism is consistently a strong predictor of positive attitudes during intergroup interactions and these positive effects emerged via perceived interaction quality. The situational needs model is robust to alternative models and outperforms Allport's contact conditions. As one of the first studies to test intergroup contact theory using extensive longitudinal data, we offer insight into the mechanisms of positive intergroup contact during real-life interactions and find situational motivations to be a key building block of understanding and addressing positive intergroup interactions.

@@

\keywords{
    Intergroup Contact, Psychological Needs, Outgroup Attitudes, Interaction Quality, Extensive Longitudinal Data\\
    \textit{Open Science Practices:}
    \noindent \href{https://osf.io}{\includegraphics[height=\badgeheight]{assets/open-badges-small/registrationplus-color.png}} Preregistration+, 
    \href{https://osf.io}{\includegraphics[height=\badgeheight]{assets/open-badges-small/material-color.png}} Open Materials, 
    \href{https://osf.io}{\includegraphics[height=\badgeheight]{assets/open-badges-small/data-color.png}} Open Data,\break
    \href{https://osf.io}{\includegraphics[height=\badgeheight]{assets/open-badges-small/code-color.png}} Open Code, 
    \href{https://osf.io}{\includegraphics[height=\badgeheight]{assets/open-badges-small/supplements-color.png}} Open Supplements
    }

@@

\section{Need Mechanism in Intergroup Contact}
% Motivational mechanism:
% interaction quality is important
% past research has either focused on conditions or cognitive-affective processes
% neither are necessary nor explain why an interaction is positive
Several meta-analytic reviews have found that positive intergroup contact reduces prejudice and increases positive attitudes in experimental and cross-sectional studies \citep[][]{Tropp2005, Pettigrew2006, Davies2011}, as well as in intergroup contact interventions outside the lab \citep[][]{Beelmann2014, Lemmer2015}. It is widely accepted that equal status, common goals, collaboration, and structural support during the interaction form the optimal conditions for such positive contact effects \citep[][]{Allport1954b, Pettigrew1969}. However, despite their prominence in guiding research on this topic, meeting the contact conditions does not seem to be essential to finding positive effects of intergroup contact \citep[][]{Pettigrew2006}. It, thus, remains unclear when and why an interaction is perceived as positive and what factors (beyond Allport's conditions) explain positive contact effects\footnote{It should be noted here that following Allport's original conditions, several additional conditions of optimal contact were proposed \citep[for a critical discussion see][]{Pettigrew1986}. Similarly, since the early 21\textsuperscript{st} century many psychological processes during intergroup contact were examined \citep[e.g. see,][]{Paolini2021}. Among others, researchers have explored different forms of social categorizations \citep[][]{Pettigrew1998}, the salience of social categories \citep[][]{Brown2005}, intimacy \citep[e.g.,][]{Marinucci2021} and attachment \citep[e.g.,][]{Tropp2021}, threat and intergroup anxiety \citep[e.g.,][]{Stephan2008, Paolini2004}, and to a lesser extent knowledge about the other group \citep[][]{Pettigrew2008c}. However, all these advances share the underlying criticism that they are not geared towards understanding when and why an interaction is perceived as positive.}. We propose that the literature on motivation and psychological needs may help us understand when and why exactly an interaction is perceived as positive.

@@

% Difficult to test general mechanism: Either consider too many needs (situationally relevant but not feasible) or too few (feasible but not transferable). Solution: ask for main need + rating of main need  
% MAYBE MOVE THIS TO 'The Present Research Section'
One reason why motivational considerations might have remained absent from the intergroup contact literature is that there is an overwhelming number of individual needs or goals that might be relevant to a person during an intergroup interaction. Researchers considering the motivational content would, thus, either test few hyper-specific needs that might not be transferable to other intergroup contexts or they may need to assess a broad range of fundamental psychological needs. To avoid this predicament we propose to start by using adaptive and responsive survey designs that allow a tailored approach based on the participants' inputs \citep[e.g.,][]{Tourangeau2017}. In particular, we propose to ask the participants to report their main goal during the interaction in a short open-ended question (i.e., situation core need) and with reference to their own response the participants can then indicate how much this need was fulfilled during the interaction (i.e., need fulfillment). Such an adaptive approach allows us to take this initial step and test whether situational psychological needs indeed generally predict perceived interaction quality and positive outgroup attitudes independent of need content.

@@

\section{The Present Research}
Using three independent sets of extensive longitudinal data (Studies 1–3), the aim of this paper is essentially threefold. First, our study is among the first to test the fundamental tenets of intergroup contact and Allport's conditions in real-life extensive longitudinal data. Translating the contact hypothesis into extensive longitudinal data is not a trivial task, as past research traditions have used two fundamentally different approaches. While lab studies have tended to focus on the effect of a single positive interaction, cross-section studies have primarily investigated the frequency of positive interactions more generally. Extensive longitudinal data allows us to investigate both. We can test whether having an interaction vs. not having an interaction impacts outgroup attitudes, but we can also use the participant's 30-day contact reports to test whether participants with more positive interactions tend to have more positive outgroup attitudes. Testing both approaches of the contact hypothesis allows us to go beyond a replication of the basic theory but could disentangle individual- from aggregated contact effects and would allow for a direct comparison with both bodies of literature. We will test the effect of individual interactions within each study using a multilevel model, but to avoid power limitations, we will test the collective effect of contact frequency and -quality after the individual studies, across all participants.
\begin{enumerate}[leftmargin=1.5\parindent]
    \item[H1:] Based on the most general understanding of the contact hypothesis, an increase in frequency and quality of contact should jointly account for more favorable outgroup attitudes within and across extensive longitudinal data.
\end{enumerate}

@@

Our second aim is to test our main theoretical proposal that the fulfillment of situational psychological needs is meaningfully related to more positive outgroup attitudes following intergroup interactions. As our main proposal is concerned with the mechanisms of successful intergroup contact, we again focus on outgroup interaction reports. Within a multi-level model, we expect interactions that are higher in situational core need fulfillment to be perceived as more positive, and as a result that these interactions also predict more positive outgroup attitudes. We also expect the needs mechanism to work at least as well as Allport's conditions. We particularly expect part of Allport's contact conditions to be a static set of situational needs so that the situational core need fulfillment should explain some of the same variance in outgroup attitudes.
\begin{enumerate}[leftmargin=1.5\parindent]
    \item[H3:] Based on our proposal, intergroup interactions with higher situational core need fulfillment should predict more favorable outgroup attitudes due to more positive interaction quality perceptions within the extensive longitudinal data. We also expect situational need fulfillments to work at least as well as Allport's optimal contact conditions in predicting outgroup attitudes.
\end{enumerate}

@@

Before turning to individual studies, two methodological matters need consideration. Firstly, we tested most of our hypotheses using multilevel regression models, where measurement occasions (level 1) were nested within participants (level 2). This approach is tolerant to missing data and uneven case numbers within participants. Furthermore, we use a hierarchical modeling approach and report the final model in-text \citep[][; for the full modeling process see Supplementary Material A]{Bliese2013}. Secondly, statistical power estimations for intensive longitudinal- and multilevel models are notoriously difficult due to the complex covariance structures. However, our sample sizes and measurement numbers are within the highest percentiles of sample sizes found within the broader intensive longitudinal literature \citep[e.g.,][]{AanhetRot2012}. Additionally, power simulations after the first study showed that our data was sufficiently powered for even small effect sizes. The sample sizes were only increased to allow for between-subjects comparisons and future, more complicated, time-series modeling.

@@

% Contact hypothesis in ESM data [inconsistent in aggregate but consistent in ML analysis]
When considering the results of the three studies jointly, we found mixed results for the basic intergroup contact hypothesis. To replicate the two common approaches to the contact hypothesis, we look at both the within-person effects of individual outgroup interactions (mimicking the analysis of lab studies) as well as the joint effect of interaction frequency and average interaction quality between participants (mimicking the cross-section literature). For the effect of individual interactions, we find that having an outgroup interaction (vs. not having an interaction) is associated with more favorable outgroup attitudes. Similarly, we find that within outgroup interactions, the interaction quality is meaningfully associated with more favorable outgroup attitudes. Yet, considering interaction frequency and average interaction quality jointly is only possible on the aggregated between-participant level. Surprisingly, here we found independent effects of interaction frequency and average interaction quality ratings but no interaction effect. The absence of this aggregate effect could underline the fact that cross-sectional retrospective data might be misleading because (1) it presents a mixture of within and between-subjects effects \citep[][]{Hamaker2020}, or (2) suffers from recall biases (e.g., where times with no outgroup interactions are undervalued by participants during the retrospective evaluations)\footnote{It should be noted that such inconsistencies with past research might in part be a data artifact (e.g., because most people reported substantially more measurements during which they did not have an outgroup interaction)}.

% Allport's conditions in ESM data [consistent predictor and partially through quality]
Using the data from our third study, we find that Allport's conditions are related to higher interaction quality perceptions and more positive outgroup attitudes. When we consider interaction quality and Allport's conditions jointly, we find that Allport's conditions also predicted whether such interactions were also perceived as more positive. We thus find first evidence that Allport's conditions of optimal contact are also relevant to the daily interactions recent migrants have in their interactions with majority group members.

% Need fulfillment in intergroup contact [consistent predictor, through quality, slightly better than Allport and SDT]
Finally, when looking at the results regarding the importance of situational key needs, we find that in all three extensive longitudinal datasets, the fulfillment of core needs during intergroup contacts predicts higher interaction quality perceptions and more positive outgroup attitudes. We find that in all three studies the effect of core need fulfillment on outgroup attitudes could either fully or partially be explained by its effect on interaction quality. We also find that need fulfillment is an important predictor even when taking other fundamental psychological needs or Allport's conditions into account. In fact, our core situational needs measure predicted outgroup attitudes slightly better than Allport's conditions and consistently explained more variance in outgroup attitudes than any of the other psychological needs. In most cases, the core need even took over the variances previously explained by the self-determination theory needs (see \appref{app:AppendixRobustness}). We thus find strong evidence that within everyday life interactions of recent migrants with majority outgroup members, the perception that one's interaction-specific psychological needs are fulfilled offers a meaningful and flexible predictor of outgroup attitudes.

@@

\subsection{Implications}
% ESM data
Despite these limitations, we can nonetheless draw a number of implications for other researchers and practitioners --- ranging from the benefits of longitudinal data to theoretical implications. A first implication concerns the feasibility and usefulness of extensive longitudinal data for intergroup contact research and the broader field of social psychology. While setting up an extensive longitudinal study is not easy, we believe the efforts to be similar to a sizable cross-sectional data collection (i.e., for a longitudinal or a high-quality cross-sectional data set with roughly 3,000 intergroup interactions captured and over 10,000 data points in total). Given the comparable effort, this opens up the possibility to explore research questions that focus on real-life phenomena outside the lab or focus on phenomena that depend on changes and influences over time. In the context of intergroup contact research, we are among the first to answer calls to test intergroup contact mechanisms using extended real-life data \citep[e.g.,][]{Pettigrew2011, MacInnis2015}. In doing so, we not only collected an unprecedented amount of real-life data, but our consideration of extensive longitudinal data may present new inconsistencies in how participants perceive and cognitively aggregate their past interactions with other groups — which may suggest large-scale recall biases or conflations of within and between participant effects in conventional cross-section studies.

% Motivation and core needs 
A broader theoretical implication relates to the role of situational motivation in intergroup contacts. Our results offer a first promising test of psychological needs in intergroup contact. While our results are tentative given their novelty within the field, they were highly consistent across studies and may offer new theoretical avenues. Psychological needs are a facet of the human experience that has thus far been under–emphasized in the intergroup contact literature. This stands in stark contrast of the many cognitive \citep[e.g.,][]{Pettigrew1998, Brown2005} and emotional aspects investigated within the field \citep[e.g.,][]{Stephan2008, Paolini2004}. Future research may, therefore, be able to integrate broader theoretical frameworks of intergroup contact (e.g., motivations guiding cognition and affect, which in turn drive behavior. cf., theory of reasoned goal pursuit; \citealp{Ajzen2019}. Also see \citealp{Kreienkamp2022d}).

% practical use of needs
Additionally, situational motivations in intergroup contact also offer promising avenues for practitioners and policy-makers. Intergroup contact theory is among the most implemented psychological theories \citep[e.g.,][]{Pettigrew2006, AlRamiah2012a, Reimer2021}. Given our findings that psychological needs in every-day intergroup contacts were at least as powerful as Allport's conditions in predicting outgroup attitudes, considerations of people's needs offer a substantially more immediate mechanism to address. In cases where some or all optimal contact conditions are not possible to be fulfilled, needs offer an even more compelling alternative (e.g., where equal status is contextually not possible or in cases where people help despite a lack of institutional support). Additionally, our conceptualization of situational needs offers a clear opportunity for practitioners and interventions. Instead of addressing needs as a one-size fits all solution (e.g., simply focusing on competence needs), one may at times ask outgroup interaction partners what they need during an interaction. This is not to say that we should not explore which needs tend to be relevant to specific groups in specific intergroup contact contexts. Rather, during interventions for which data on important need contents are not available or infeasible to collect, a flexible and reactive approach of inquiring momentary intergroup contact needs might be more fruitful.

>>>>

% Abstract
\abstract{
One challenge of modern intergroup contact research has been the question of when and why an interaction is perceived as positive and leads to better intergroup relations. We propose to consider situational psychological needs during everyday intergroup contact. We conducted three extensive longitudinal studies with recent migrants, to capture their interactions with the majority outgroup (total \textit{N} of measurements = 10,297). The novel psychological needs mechanism is consistently a strong predictor of positive attitudes during intergroup interactions and these positive effects emerged via perceived interaction quality. The situational needs model is robust to alternative models and outperforms Allport's contact conditions. As one of the first studies to test intergroup contact theory using extensive longitudinal data, we offer insight into the mechanisms of positive intergroup contact during real-life interactions and find situational motivations to be a key building block of understanding and addressing positive intergroup interactions.

@@

\keywords{
    Intergroup Contact, Psychological Needs, Outgroup Attitudes, Interaction Quality, Extensive Longitudinal Data\\
    \textit{Open Science Practices:}
    \noindent \href{https://osf.io/r5e8c?view_only=bfaed2196d49420385ce5fa6506a7765}{\includegraphics[height=\badgeheight]{assets/open-badges-small/registrationplus-color.png}} Preregistration+, 
    \href{https://osf.io/pr9zs/?view_only=1ea47bb646694632a764dead807ef970}{\includegraphics[height=\badgeheight]{assets/open-badges-small/material-color.png}} Open Materials, 
    \href{https://osf.io/pr9zs/?view_only=1ea47bb646694632a764dead807ef970}{\includegraphics[height=\badgeheight]{assets/open-badges-small/data-color.png}} Open Data,\break
    \href{https://osf.io/pr9zs/?view_only=1ea47bb646694632a764dead807ef970}{\includegraphics[height=\badgeheight]{assets/open-badges-small/code-color.png}} Open Code, 
    \href{https://osf.io/pr9zs/?view_only=1ea47bb646694632a764dead807ef970}{\includegraphics[height=\badgeheight]{assets/open-badges-small/supplements-color.png}} Open Supplements
    }

@@

\section{Need Mechanism in Intergroup Contact}
% Motivational mechanism:
% interaction quality is important
% past research has either focused on conditions or cognitive-affective processes
% neither are necessary nor explain why an interaction is positive
Several meta-analytic reviews have found that positive intergroup contact reduces prejudice and increases positive attitudes in experimental and cross-sectional studies \citep[][]{Tropp2005, Pettigrew2006, Davies2011}, as well as in intergroup contact interventions outside the lab \citep[][]{Beelmann2014, Lemmer2015}. It is widely accepted that equal status, common goals, collaboration, and structural support during the interaction form the optimal conditions for such positive contact effects \citep[][]{Allport1954b, Pettigrew1969}. However, despite their prominence in guiding research on this topic, meeting the contact conditions does not seem to be essential to finding positive effects of intergroup contact \citep[][]{Pettigrew2006}. It, thus, remains unclear when and why an interaction is perceived as positive and what factors (beyond Allport's conditions) explain positive contact effects\footnote{It should be noted here that following Allport's original conditions, several additional conditions of optimal contact were proposed \citep[for a critical discussion see][]{Pettigrew1986}. Similarly, since the early 21\textsuperscript{st} century many psychological processes during intergroup contact were examined \citep[e.g. see,][]{Paolini2021}. Among others, researchers have explored different forms of social categorizations \citep[][]{Pettigrew1998}, the salience of social categories \citep[][]{Brown2005}, intimacy \citep[e.g.,][]{Marinucci2021} and attachment \citep[e.g.,][]{Tropp2021}, threat and intergroup anxiety \citep[e.g.,][]{Stephan2008, Paolini2004}, and knowledge about the other group \citep[][]{Pettigrew2008c}. However, all these advances share the underlying criticism that they are not geared towards understanding when and why an interaction is perceived as positive.}. We propose that the literature on motivation and psychological needs may help us understand when and why exactly an interaction is perceived as positive.

@@

% Difficult to test general mechanism: Either consider too many needs (situationally relevant but not feasible) or too few (feasible but not transferable). Solution: ask for main need + rating of main need  
% MAYBE MOVE THIS TO 'The Present Research Section'
One reason why motivational considerations might have remained absent from the intergroup contact literature is that there is an overwhelming number of individual needs or goals that might be relevant to a person during an intergroup interaction. Researchers considering the motivational content would, thus, either test few hyper-specific needs that might not be transferable to other intergroup contexts or they may need to assess a broad range of fundamental psychological needs. To avoid this predicament we propose to start by using adaptive and responsive survey designs that allow a tailored approach based on the participants' inputs \citep[e.g.,][]{Tourangeau2017}. In particular, we propose to ask the participants to report their main goal during the interaction in a short open-ended question (i.e., situation core need) and with reference to their own response, the participants can then indicate how much this need was fulfilled during the interaction (i.e., need fulfillment). Such an adaptive approach allows us to take the initial step of testing whether situational psychological needs indeed generally predict perceived interaction quality and positive outgroup attitudes independent of need content.

@@

\section{The Present Research}
Using three independent sets of extensive longitudinal data (Studies 1–3), the aim of this paper is essentially threefold. First, our study is among the first to test the fundamental tenets of intergroup contact and Allport's conditions in real-life extensive longitudinal data. Translating the contact hypothesis into extensive longitudinal data is not a trivial task, as past research traditions have used two fundamentally different approaches. While lab studies have tended to focus on the effect of a single positive interaction, cross-section studies have primarily investigated the frequency of positive interactions more generally. Extensive longitudinal data allows us to investigate both. We can test whether having an interaction vs. not having an interaction impacts outgroup attitudes, but we can also use the participant's 30-day contact reports to test whether participants with more positive interactions tend to have more positive outgroup attitudes. Testing both approaches to the contact hypothesis allows us to go beyond a replication of the basic theory but could disentangle individual- from aggregated contact effects and would allow for a direct comparison with both bodies of literature. We will test the effect of individual interactions within each study using a multilevel model, but to avoid power limitations, we will test the collective effect of contact frequency and -quality after the individual studies, across all participants.
\begin{enumerate}[leftmargin=1.5\parindent]
    \item[H1:] Based on the most general understanding of the contact hypothesis, an increase in frequency and quality of contact should jointly account for more favorable outgroup attitudes within and across extensive longitudinal data.
\end{enumerate}

@@

Our second main aim is to test our main theoretical proposal that the fulfillment of situational psychological needs is meaningfully related to more positive outgroup attitudes following intergroup interactions. As our main proposal is concerned with the mechanisms of successful intergroup contact, we again focus on outgroup interaction reports. Within a multi-level model, we expect interactions that are higher in situational core need fulfillment to be perceived as more positive, and as a result that these interactions also predict more positive outgroup attitudes. We also expect the needs mechanism to work at least as well as Allport's conditions. We particularly expect part of Allport's contact conditions to be a static set of situational needs so that the situational core need fulfillment should explain some of the same variance in outgroup attitudes.
\begin{enumerate}[leftmargin=1.5\parindent]
    \item[H3:] Based on our proposal, intergroup interactions with higher situational core need fulfillment should predict more favorable outgroup attitudes due to more positive interaction quality perceptions within the extensive longitudinal data. We also expect situational need fulfillments to work at least as well as Allport's optimal contact conditions in predicting outgroup attitudes.
\end{enumerate}

@@

Before turning to individual studies, two methodological matters need consideration. Firstly, we tested most of our hypotheses using multilevel regression models, where measurement occasions (level 1) were nested within participants (level 2). This approach is tolerant to missing data and uneven case numbers within participants. Furthermore, we use a hierarchical modeling approach and report the final model in-text \citep[][; for the full modeling process see Supplementary Material A]{Snijders2012}. Secondly, statistical power estimations for intensive longitudinal- and multilevel models are notoriously difficult due to the complex covariance structures. However, our participant- and measurement numbers are within the highest percentiles of sample sizes found within the broader intensive longitudinal literature \citep[e.g.,][]{AanhetRot2012}. Additionally, power simulations after the first study showed that our data was sufficiently powered for even small effect sizes. The sample sizes were only increased to allow for between-subjects comparisons and future, more complicated, time-series modeling.

@@

% Contact hypothesis in ESM data [inconsistent in aggregate but consistent in ML analysis]
When considering the results of the three studies jointly, we found mixed results for the basic intergroup contact hypothesis. To replicate the two common approaches to the contact hypothesis, we looked at both the within-person effects of individual outgroup interactions (mimicking the analysis of lab studies) as well as the joint effect of interaction frequency and average interaction quality between participants (mimicking the cross-section literature). For the effect of individual interactions, we find that having an outgroup interaction (vs. not having an interaction) was associated with more favorable outgroup attitudes. Similarly, we find that within outgroup interactions, the interaction quality was meaningfully associated with more favorable outgroup attitudes. Yet, considering interaction frequency and average interaction quality jointly was only possible on the aggregated between-participant level. Surprisingly, here we found independent effects of interaction frequency and average interaction quality ratings but no interaction effect. The absence of this aggregate effect could underline the fact that cross-sectional retrospective data might be misleading because (1) it presents a mixture of within and between-subjects effects \citep[][]{Hamaker2020}, or (2) suffers from recall biases (e.g., where times with no outgroup interactions are undervalued by participants during the retrospective evaluations)\footnote{It should be noted that such inconsistencies with past research might in part be a data artifact (e.g., because most people reported substantially more measurements during which they did not have an outgroup interaction).}.

% Allport's conditions in ESM data [consistent predictor and partially through quality]
Using the data from our third study, we find that Allport's conditions are related to higher interaction quality perceptions and more positive outgroup attitudes. When we consider interaction quality and Allport's conditions jointly, we find that the positive effect of Allport's conditions might in part be explained by more positive interaction quality ratings. We thus find first evidence that Allport's conditions of optimal contact are also relevant to the daily interactions recent migrants have in their interactions with majority group members.

% Need fulfillment in intergroup contact [consistent predictor, through quality, slightly better than Allport and SDT]
Finally, when looking at the results regarding the importance of situational key needs, we find that in all three extensive longitudinal data sets, the fulfillment of core needs during intergroup contacts predicts higher interaction quality perceptions and more positive outgroup attitudes. We find that in all three studies the effect of core need fulfillment on outgroup attitudes could either fully or partially be explained by its effect on interaction quality. We also find that need fulfillment is an important predictor even when taking other fundamental psychological needs or Allport's conditions into account. In fact, our core situational needs measure predicted outgroup attitudes slightly better than Allport's conditions and consistently explained more variance than commonly measured psychological needs. In most cases, the core need even took over the variances previously explained by the self-determination theory needs (see \appref{app:AppendixRobustness}). We thus find strong evidence that within everyday life interactions of recent migrants with majority outgroup members, the perception that one's interaction-specific psychological needs are fulfilled offers a meaningful and flexible predictor of outgroup attitudes.

@@

\subsection{Implications}
% ESM data
Despite these limitations, we can nonetheless draw a number of implications for other researchers and practitioners --- ranging from the benefits of longitudinal data to theoretical implications. A first implication concerns the feasibility and usefulness of extensive longitudinal data for intergroup contact research and the broader field of social psychology. While setting up an extensive longitudinal study is not easy, we believe the efforts to be similar to a sizable cross-sectional data collection (i.e., for a longitudinal or a high-quality cross-sectional data set with roughly 3,000 intergroup interactions captured and over 10,000 data points in total). Intensive longitudinal data, thus, opens up the possibility to explore research questions that focus on real-life phenomena outside the lab or focus on phenomena that depend on changes and influences over time. In the context of intergroup contact research, we are among the first to answer calls to test intergroup contact mechanisms using extended real-life data \citep[e.g.,][]{Pettigrew2011, MacInnis2015}. In doing so, we not only collected an unprecedented amount of real-life data, but our consideration of extensive longitudinal data may present new inconsistencies in how participants perceive and cognitively aggregate their past interactions with other groups — which may suggest large-scale recall biases or conflations of within and between participant effects in conventional cross-section studies.

% Motivation and core needs 
A broader theoretical implication relates to the role of situational motivation in intergroup contacts. Our results offer a first promising test of psychological needs in intergroup contact. While our results are tentative given their novelty within the field, they were highly consistent across studies and may offer new theoretical avenues. Psychological needs are a facet of the human experience that has thus far been underemphasized in the intergroup contact literature. This stands in stark contrast of the many cognitive \citep[e.g.,][]{Pettigrew1998, Brown2005} and emotional aspects investigated within the field \citep[e.g.,][]{Stephan2008, Paolini2004}. Future research may, therefore, be able to integrate broader theoretical frameworks of intergroup contact (e.g., motivations guiding cognition and affect, which in turn drive behavior. cf., theory of reasoned goal pursuit; \citealp{Ajzen2019}. Also see \citealp{Kreienkamp2022d}).

% practical use of needs
Additionally, situational motivations in intergroup contact also offer promising avenues for practitioners and policy-makers. Intergroup contact theory is among the most implemented psychological theories \citep[e.g.,][]{Pettigrew2006, AlRamiah2012a, Reimer2021}. Given our findings that psychological needs in everyday intergroup contacts were at least as powerful as Allport's conditions in predicting outgroup attitudes, considerations of people's needs offer a substantially more immediate mechanism to address. In cases where some or all optimal contact conditions are not possible to be fulfilled, needs offer an even more compelling alternative (e.g., where equal status is contextually not possible or in cases where people help despite a lack of institutional support). Additionally, our conceptualization of situational needs offers a clear opportunity for practitioners and interventions. Instead of addressing needs as a one-size fits all solution (e.g., simply focusing on competence needs), one may at times ask outgroup interaction partners what they need during an interaction. This is not to say that we should not explore which needs tend to be relevant to specific groups in specific intergroup contact contexts. Rather, during interventions for which data on important need contents are not available or infeasible to collect, a flexible and reactive approach of inquiring momentary intergroup contact needs might be more fruitful.
